Increasing Performance via Gamification in a Volunteer-Based Evolutionary Computation System

### 2018-03-08 2018-IPMU/Gamification/paper.tex
REcoloca correos
Edit to address some the comments from  reviewers.
Edit to address some the comments from  reviewers.  Added a new
proposal section.

\institute{Instituto Tecnol\'ogico de Tijuana, Tijuana BC, Mexico
\and
Universidad de Granada, Granada, Spain
\and
Universidad de C\'adiz, C\'adiz, Spain
\email{mario@tectijuana.edu.mx}\\
\email{jmerelo@geneura.ugr.es}}

@@

\section{Introduction}
The world wide web is an increasingly reliable and high-performance
operating system for running distributed applications. Besides the
maturity of HTTP, the underlying protocol itself, there are other
factors: the JavaScript virtual machine embedded in every browser
and the REST interface, providing a {\em de facto} standard interface for
communication. A distributed system based on the browser can be
easily created using these mechanisms, which lately have been enhanced
by WebSockets for lower latency; you can start to gather users just by
announcing the URL. This approach for creating distributed
experiments is called {\em volunteer}, {\em cycle-scavenging},
or {\em opportunistic} computing
\cite{sarmenta2001volunteer} and it dates back, in different shapes
and underlying mechanisms, to the origin of the Internet
\cite{david-seti:home}.

@@

But even if adaptation is straightforward, the application of volunteer
computing techniques to the development of distributed evolutionary algorithms
still has several open issues. An important research aspect is
approaching the problem as a socio-technical system
\cite{vespignani2009predicting,merelo2015designing}, which integrates
user decisions and behavioral patterns in the model; this
includes trying to optimize the number of users or the overall time
they contribute in a particular
experiment. The challenge is to design a system that, whatever the
number of users available and willing to perform the experiment, it is able to
maximize their contribution to the evolutionary algorithm. One obvious
way of achieving that is trying to find as many users as possible. But
another approach, and the one that we will be following in this paper, is
to delve into the socio-technical nature of the system, using
gamification techniques to try and improve the amount of time every user will
lend to the system. However, there is a trade-off with gamification,
since users have to be identified in some way to participate, and thus
it has to include some way of user authentication,
which is a hurdle that might take users away; however, the open design %Pablo: use colon instead semicolon?
of the system we use means that we can actually mix anonymous (with no
gamification) and authenticated (gamified) users, with only the small overhead
of adding the gamification mechanism.

@@

The rest of the paper is organized as follows: The state of the art in opportunistic/volunteer distributed evolutionary
computation (EC) and Gamification Techinques as presented next. Section \ref{sec:experiments} will describe the
framework and the problem used in the experiments, which are publicly
available under a free license. The results of each of the steps in the incremental design are presented in Section
\label{sec:results}, to finally wrap up with the conclusions. %Pablo: check out this paragraph! Section 3 is not describing the problem and framework, it is only discussing on related work on gamification! I suggest combine section Gamification into Section 2, and separate Section 4.2 (Results) as a new Section before Conclusions.
% Done - Mario

@@

Many systems described above do not go any further than trying to find out how many users join the effort, and how many of them the system can support. In fact, systems such as the one described in
\cite{gecco07:workshop:dcor} had severe scaling issues; some of them
also tried to find out how much time was needed to find the solution
or, alternatively, how many users would be needed to be competitive
against single-user single-computer implementations of the same
algorithm. Lately, researchers have tried to integrate volunteer
computing techniques as a part of a larger distributed evolutionary computing
effort \cite{leclerc2016seamless}.  In fact, systems using volunteers
exclusively, exhibit a certain amount of unpredictability, and they
might be better used in combination with ready-available computing
power.

\section{Gamification}
\label{sec:gamification}
A definition given by Huotari \cite{huotari2012defining} is ``Gamification is
the process by which gaming concepts are brought to real world tasks associated with
real people''. Gamification uses game design elements out of the domain of games
with the objective of enhancing the user\'s experience, engagement, productivity,
learning, among others. Deterding et al. proposes the following definition:
 ``Gamification” is the use of game design elements in non-game contexts''
 \cite{deterding2011gamification}.

Gamification techniques in a volunteer context seeks to persuade
users to use their natural desire to compete, learn and socialize in
given non-game context application \cite{deterding2011game,hamari2014does}.
% Some works give a form of reward to users, these include
% points \cite{sutter2010browse}, achievement badges or levels \cite{hamari2011framework},
% the filling of a progress bar \cite{o2010get}, or providing the user with virtual currency.
By Making the rewards for  tasks achievements visible to other players or
providing leader boards are ways of encouraging players to compete
\cite{hickman2010total}. In this work, a leader board was implemented
in order to promote competition, presenting only the all-time top five
participating users.
% Competition can also have problematic consequences, which can result in
% negative conduct, low cooperation and collaboration, or disadvantaging certain player demographics
% such as women \cite{kumar2013gamification}.

@@

In this experiment a gamification technique is applied to
the {\sf NodIO}
volunteer computing framework, using the
particular version described in \cite{DBLP:conf/gecco/MereloCGCRV16,2016arXiv160101607M}.
The objective of the experiment is to test the kind of impact
applying a gamification technique has on user engagement in this
particular application. In order to apply a rewarding system
user authentication had to be developed first. In earlier designs
this functionality was not desired because it was seen as a barrier
for participation. After all, the advantage of a browser based volunteer system
is precisely the minimum amount of user intervention needed to start.
The last thing a user wants to see is yet another registration form.
Then, the first design decision for this version is that registration
is optional and simple.
%as shown in Figure \ref{fig:login}.

The gamification technique employed in this work is based on a rewarding mechanism
\cite{dubois2013understanding}. In general rewards  consist of a reputation system
with score points, levels and leader boards. Points are awarded to users in response of
the accomplishment of certain activities that need to be encouraged. In these case
a point is awarded for each HTTP {\tt PUT} Request sent to the server. Levels are a long
term achievement, in this case the level depends on the score:
\[ \text{level}(\text{score};a,b)=
    \begin{cases}
      0,                                    &  \text{score}\leq a\\
      2(\frac{\text{score}-a}{b-a})^{2},    &  a\leq \text{score}\leq \frac{a+b}{2}\\
      1-2(\frac{\text{score}-b}{b-a})^{2},  & \frac{a+b}{2} \leq \text{score}\leq b\\
      1,                                    & \text{score}\geq b
   \end{cases}
\]
The function returns a normalized value that is multiplied by the maximum level
in this case 100. The variables $a$=100 and $b$=7360 are set to give users a rapid increase of
levels at the beginning.%; this function is represented in Figure \ref{fig:s}.
%
% \begin{figure*}[htb]
%     \centering
%         \includegraphics[width=3in]{img/s.png}
%     \caption{$S$ function used to assign the user's level.
%     }
%     \label{fig:s}
% \end{figure*}

After login to the NodIO web application, %Pablo: after login where? Add a link to the page or explain that the system was run in our servers
registered users are welcomed by their name, and they can see
their current score and level. All users can see the
leader board in a modal dialog.
While the modal is opened, the scores are refreshed every second. All other
functionality is available to all users and is the same as the
previous {\sf NodIO} version, showing the state of the current algorithm.
In the front page a link to the open source code of the experiment was also
available \url{https://github.com/lucero21/login-master}. %Pablo: I suggest to move the link to the introduction, just after the first time it is mentioned as free software

@@

\subsection{Results}
\label{sec:results}

@@

The second observation is apparent in the $y$ axis of Figure
\ref{fig:puts}. There is a notable difference in the amount of
participation between the two groups, and also in the slope of the
plot.  An independent-samples t-test was conducted to compare
participation, and there was a significant difference in the scores
for anonymous users  (M=1057.97, SD= 3669.86) and non-anonymous users
(M=46513.55, SD=93537.20); t-test= 4.70, p = 7.5e-06; that is, the
contribution of registered users was two orders of magnitude bigger
than for anonymous users. This difference is highlighted in
%
%Maybe a statistical comparison? Median, for instance - JJ
%count          mean           std  min      25%      50%      75%
%category
%IP        91.0   1057.978022   3669.862398  1.0     1.50     18.0    173.0
%user      18.0  46513.555556  93537.207948  7.0  3962.75  13034.5  46993.5
%
%
%      max
%category
%IP         26593.0
%user      396081.0
%Ttest_indResult(statistic=4.7070787365897289, pvalue=7.565505602664807e-06)
%
%
\begin{figure*}[htb]
    \centering
        \includegraphics[width=5in]{img/puts_box.png}
    \caption{Box-plot of the number of requests sent by registered users and anonymous IPs.
     In the \emph{y} axis are number or PUTs in a logarithmic scale.
    }
    \label{fig:box}
\end{figure*}
%
 Figure \ref{fig:box}, which is a boxplot of the  participation by two
 groups, shows the big difference between registered users (shown as
 ``users'') and anonymous ones (shown as ``IPs''). This chart also
 shows the number of outliers in the groups, which is an obvious
 consequence of the ephemeral and spontaneous nature of the
 collaboration; in fact, two registered
outliers were very important participants, maybe competing between
them, an expected consequence of the gamification, in fact, although
not in this precise form and quantity. But
there were more outliers between anonymous IPs, and some of them could
be related to the same users. The lower quartile in the users box-plot
has more
spread, indicating a participation similar to the median of anonymous
IPs, indicating that the difference in distribution is mainly due to a
few highly engaged and registered users.

>>>>

\institute{Instituto Tecnol\'ogico de Tijuana, Tijuana BC, Mexico; 
  \email{mario@tectijuana.edu.mx}
\and
Universidad de Granada, Granada, Spain; 
\email{jmerelo@geneura.ugr.es}
\and
Universidad de C\'adiz, C\'adiz, Spain
}

@@

\section{Introduction}
The World Wide Web is an increasingly reliable and high-performance
operating system for running distributed applications. Besides the
maturity of HTTP, the underlying protocol itself, there are other
factors: the JavaScript virtual machine embedded in every browser
and the REST interface, providing a {\em de facto} standard interface for
communication. A distributed system based on the browser can be
easily created using these mechanisms, which lately have been enhanced
by WebSockets for lower latency; you can start to gather users just by
announcing the URL. This approach for creating distributed
experiments is called {\em volunteer}, {\em cycle-scavenging},
or {\em opportunistic} computing
\cite{sarmenta2001volunteer} and it dates back, in different shapes
and underlying mechanisms, to the origin of the Internet
\cite{david-seti:home}.

@@

But even if adaptation is straightforward, the application of volunteer
computing techniques to the development of distributed evolutionary algorithms
still has several open issues. An important research aspect is
approaching the problem as a socio-technical system
\cite{vespignani2009predicting,merelo2015designing}, which integrates
user decisions and behavioral patterns in the model; this
includes trying to optimize the number of users or the overall time
they contribute in a particular
experiment. The challenge is to design a system that, whatever the
number of users available and willing to perform the experiment, it is able to
maximize their contribution to the evolutionary algorithm. One obvious
way of achieving that is trying to find as many users as possible. But
another approach, and the one that we will be following in this paper, is
to delve into the socio-technical nature of the system, using
gamification techniques to try and improve the amount of time every user will
lend to the system. However, there is a trade-off with gamification,
since users have to be identified in some way to participate, and thus
it has to include some way of user authentication,
which is a hurdle that might take users away. However, the open design %Pablo: use colon instead semicolon?
of the system we use means that we can actually mix anonymous (with no
gamification) and authenticated (gamified) users, with only the small overhead
of adding the gamification mechanism.

@@

The rest of the paper is organized as follows: The state of the art in opportunistic/volunteer distributed evolutionary
computation (EC) is presented next. The proposed Gamification Techinque is described in 
Section~\ref{sec:gamification}. Section~\ref{sec:experiments} will describe the
framework and the problem used in the experiments, which are publicly
available under a free license. The results of each of the steps in the incremental design are presented in Section~\ref{sec:results}, to finally wrap up with the conclusions. 
%Pablo: check out this paragraph! Section 3 is not describing the problem and framework, it is only discussing on related work on gamification! I suggest combine section Gamification into Section 2, and separate Section 4.2 (Results) as a new Section before Conclusions.
% Done - Mario

@@

Many of the systems described above do not go any further than trying to find out how many users join the effort, and how many of them the system can support. In fact, systems such as the one described in
\cite{gecco07:workshop:dcor} had severe scaling issues; some of them
also tried to find out how much time was needed to find the solution
or, alternatively, how many users would be needed to be competitive
against single-user single-computer implementations of the same
algorithm. Lately, researchers have tried to integrate volunteer
computing techniques as a part of a larger distributed evolutionary computing
effort \cite{leclerc2016seamless}.  In fact, systems using volunteers
exclusively, exhibit a certain amount of unpredictability, and they
might be better used in combination with ready-available computing
power.

\section{Proposed Gamification Technique}
\label{sec:gamification}
A definition of Gamification given by Huotari \cite{huotari2012defining} is ``Gamification is
the process by which gaming concepts are brought to real world tasks associated with
real people''. Gamification uses game design elements out of the domain of games
with the objective of enhancing the user\'s experience, engagement, productivity,
learning, among others. Deterding et al. proposes the following definition:
 ``Gamification” is the use of game design elements in non-game contexts''
 \cite{deterding2011gamification}.

Gamification techniques in a volunteer context seeks to persuade
users to use their natural desire to compete, learn and socialize in
given non-game context application \cite{deterding2011game,hamari2014does}.
% Some works give a form of reward to users, these include
% points \cite{sutter2010browse}, achievement badges or levels \cite{hamari2011framework},
% the filling of a progress bar \cite{o2010get}, or providing the user with virtual currency.
By Making the rewards for  tasks achievements visible to other players or
providing leader boards are ways of encouraging players to compete
\cite{hickman2010total}.

In this work, a leader board was implemented
in order to promote competition, presenting only the all-time top five
participating users.

Another gamification technique employed in this work is based on a rewarding mechanism
\cite{dubois2013understanding}. In general rewards  consist of a reputation system
with score points, levels and leader boards. Points are awarded to users in response of
the accomplishment of certain activities that need to be encouraged. In these case
a point is awarded for each HTTP {\tt PUT} Request sent to the server. Levels are a long
term achievement, in this case the level depends on the score:
\[ \text{level}(\text{score};a,b)=
    \begin{cases}
      0,                                    &  \text{score}\leq a\\
      2(\frac{\text{score}-a}{b-a})^{2},    &  a\leq \text{score}\leq \frac{a+b}{2}\\
      1-2(\frac{\text{score}-b}{b-a})^{2},  & \frac{a+b}{2} \leq \text{score}\leq b\\
      1,                                    & \text{score}\geq b
   \end{cases}
\]
The function returns a normalized value that is multiplied by the maximum level
in this case 100. The variables $a$=100 and $b$=7360 are set to give users a rapid increase of
levels at the beginning. The effect $a$ and $b$ have on the results was not 
determined, and other techinques could be used.

 %; this function is represented in Figure \ref{fig:s}.

% Competition can also have problematic consequences, which can result in
% negative conduct, low cooperation and collaboration, or disadvantaging certain player demographics
% such as women \cite{kumar2013gamification}.

@@

For this experiment the gamification technique is applied to
the {\sf NodIO} volunteer computing framework, using the
particular version described in detail in \cite{DBLP:conf/gecco/MereloCGCRV16,2016arXiv160101607M}.
The objective of the experiment is to test the kind of impact
applying a gamification technique has on user engagement in this
particular application. In order to apply a rewarding system
user authentication had to be developed first. In earlier designs
this functionality was not desired because it was seen as a barrier
for participation. After all, the advantage of a browser based volunteer system
is precisely the minimum amount of user intervention needed to start.
The last thing a user wants to see is yet another registration form.
Then, the first design decision for this version is that registration
is optional and simple.
%as shown in Figure \ref{fig:login}.

%
% \begin{figure*}[htb]
%     \centering
%         \includegraphics[width=3in]{img/s.png}
%     \caption{$S$ function used to assign the user's level.
%     }
%     \label{fig:s}
% \end{figure*}

After login to the NodIO web application, %Pablo: after login where? Add a link to the page or explain that the system was run in our servers
registered users are welcomed by their name, and they can see
their current score and level. All users can see the
leader board in a modal dialog.
While the modal is opened, the scores are refreshed every second. All other
functionality is available to all users and is the same as the
previous {\sf NodIO} version, showing the state of the current algorithm.
In the front page a link to the open source code of the experiment was also
available in \url{https://github.com/lucero21/login-master}. %Pablo: I suggest to move the link to the introduction, just after the first time it is mentioned as free software

@@

\section{Results}
\label{sec:results}

@@

The second observation is apparent in the $y$ axis of Figure~\ref{fig:puts}.
There is a notable difference in the amount of
participation between the two groups, and also in the slope of the
plot.  An independent-samples t-test was conducted to compare
participation, and there was a significant difference in the scores
for anonymous users  (M=1057.97, SD= 3669.86) and non-anonymous users
(M=46513.55, SD=93537.20); t-test= 4.70, p = 7.5e-06; that is, the
contribution of registered users was two orders of magnitude bigger
than for anonymous users. This difference is highlighted in
%
%Maybe a statistical comparison? Median, for instance - JJ
%count          mean           std  min      25%      50%      75%
%category
%IP        91.0   1057.978022   3669.862398  1.0     1.50     18.0    173.0
%user      18.0  46513.555556  93537.207948  7.0  3962.75  13034.5  46993.5
%
%
%      max
%category
%IP         26593.0
%user      396081.0
%Ttest_indResult(statistic=4.7070787365897289, pvalue=7.565505602664807e-06)
%
%
\begin{figure*}[htb]
    \centering
        \includegraphics[width=5in]{img/puts_box.png}
    \caption{Box-plot of the number of requests sent by registered users and anonymous IPs.
     In the \emph{y} axis are number or PUTs in a logarithmic scale.
    }
    \label{fig:box}
\end{figure*}
%
 Figure~\ref{fig:box}, which is a boxplot of the  participation by two
 groups, shows the big difference between registered users (shown as
 ``users'') and anonymous ones (shown as ``IPs''). This chart also
 shows the number of outliers in the groups, which is an obvious
 consequence of the ephemeral and spontaneous nature of the
 collaboration; in fact, two registered
outliers were very important participants, maybe competing between
them, an expected consequence of the gamification, in fact, although
not in this precise form and quantity. But
there were more outliers between anonymous IPs, and some of them could
be related to the same users. The lower quartile in the users box-plot
has more
spread, indicating a participation similar to the median of anonymous
IPs, indicating that the difference in distribution is mainly due to a
few highly engaged and registered users.
